
```{r}

"Hello World!!"

```


# Arithemtic operation
R offers various arithmetic capabilities, including basic arithmetic operations 
(addition, subtraction, multiplication, division)

## Addition (+)
Addition is the arithmetic operation used to combine two or more numbers to find their sum. The addition operator in R is denoted by the plus sign (+).

```{r}

1 + 1

20 + 40

```


## Subtraction (-)
Subtraction is the arithmetic operation used to find the difference between two numbers. It represents the process of taking away one quantity from another. The subtraction operator in R is denoted by the minus sign (-).


```{r}

4 - 2

100 - 50

```

## Division (/)
Division is the arithmetic operation used to distribute a quantity into equal parts or to find the quotient when one number is divided by another. The division operator in R is denoted by the forward slash (/)
```{r}

100 / 2

0.5 / 0.25
```


## Multiplication (*)
Multiplication is the arithmetic operation used to find the product of two or more numbers. It represents the process of repeated addition. The multiplication operator in R is denoted by the asterisk (*)
```{r}

3 * 5

```

