# Assignment Operator
In R, assignment operators are used to store values or expressions into variables. They are crucial for creating and updating variables, which are essential for data manipulation and computations in R.


## Leftward Assignment ( <- )
Also Known as the regular assignment operator, represented by `<-`, assigns the value on its right-hand side to the variable on its left-hand side.


```{r}

x <- 10


```

The code `x <- 10` assigns the value 10 to the variable `x`. This means that the variable `x` now holds the value 10 and can it can be used for further calculations or operations in R. 

```{r}

y <- "Hello World!"

y

```


## Rightward Assignment ( -> )
The rightward assignment operator, assigns the value on its left-hand side to the value on its right-hand side.


```{r}

10 -> a

"Hello World!" -> b

a
b
```


## Assigning Results of Expressions
We can perform several calculations with the value assigned to a variable, for example the code below multiply 2 by the value assigned to `x` which is 10, then it add the value assigned to the variable `a` which is 20.

```{r}

x <- 10
a <- 20

result <- 2 * x + a  

result

```


## Overwriting Variable Values
The value assigned to a variable can be overwriting by assigning a new value to the variable. For Example the code below assign 12 to `x`  then overwrite the value by assigning 15 to the same variable. When you run the code or print(x), the output will be 15
```{r}

x <- 12
x

x <- 15
x

```


## Chain assignment
You can also assign multiple value at once by chaining the assignment operator to several variables.

```{r}

a <- x <- 20

a
x
```





