# data structure


## Vector
A vector is a fundamental data structure used to store a collection of elements of the `same data type`. It can be thought of as a one-dimensional array that holds values and those values can be numeric, character, logical or other types.

creating Vectors:

```{r}

# Numeric vector
age <- c(15, 25)

age

# Character vector
gender <- c("Female", "male")

gender


# Logical vector
logical_vector <- c(TRUE, FALSE)

logical_vector
  
```


### Named vector
A named vector consist of a collection of elements with each element having a name or an identifier associated with it.


```{r}
name <- c("Elizabeth", "John")

names(age) <- name

age
```


```{r}
names(gender) <- name

gender
```



### Coercion 
The process of coercion involves converting or transforming data from one type to another to ensure consistency within the vector.

If you combine numeric data with character data, R will convert all the elements into character data


```{r}

c(1, "A", 20.4)       

```



If you mix logical data (TRUE/FALSE) with numeric data, R will convert all the elements into numeric data (TRUE -> 1,  FALSE -> 0)

```{r}

c(TRUE, 2, FALSE, 4)

```


If you combine different numeric types (e.g, integer and double), R will convert all the element to the more general data type (double)

```{r}
c(1L, 2.0, 3L, 4.1)
```


### Combine multiple vectors
  R can combine multiple vectors using the `c()` function, so it is important you understand the underlying data type to avoid coercion.

```{r}

name <- c("Elizabeth", "John")

gender <- c("Female", "male")

c(name, gender)

```


### Vectorization
Vectorization is a powerful concept in R, where operations are applied to entire vectors instead of individual elements. When using vectorized operations, R automatically loops through each element of the vectors and performs the operation efficiently.


```{r}

salary <- c(1000, 5000, 2000)

salary * 2

salary / 2

```


```{r}

bonus <- c(500, 2500, 1000)  

salary + bonus

```




## Matrix
In R, a matrix is a two-dimensional data structure that consists of rows and column, where each element in the matrix is of the same data type.

Creating a Matrix
you can create a matrix if R using the `matrix()` function. The matrix() function takes the date elements.

```{r}

vec <- c(1, 2, 3, 4, 5, 6)

mat <- matrix(
  vec,
  nrow = 2,
  ncol = 3
)

mat

```


### using cbind
Alternatively, you can also create a matrix using the cbind() function which bind vectors column-wise

```{r}

vec1 <- c(1, 2, 3)
vec2 <- c(4, 5, 6)

mat <- cbind(vec1, vec2)

mat

```


### using rbind
You can also create a matrix using the rbind() function which binds vectors row-wise

```{r}

vec1 <- c(1, 2, 3)
vec2 <- c(4, 5, 6)

mat <- rbind(vec1, vec2)

mat

```


### Accessing Elements in a matrix
You can access elements in a matrix using row and column indices inside square brackets.

```{r}

mat <- matrix(
  1:9,
  nrow = 3,
  ncol = 3
)

mat

```


Get the elements in the second row and third column.
```{r}

mat[2, 3] 

```


Get the elements in the second & third row and also first column.
```{r}

mat[2:3, 1]

```


Get the elements in the second & third row and the first & second column.
```{r}

mat[2:3, 1:2]

```


### Matrix Operations
Matrices in R support various mathematical operations, such as addition, subtraction, multiplication, and transposition.


```{r}

mat1 <- matrix(1:4, nrow = 2)
mat2 <- matrix(5:8, nrow = 2)

mat1
mat2

```


Addition
```{r}

mat1 + mat2

```


Multiplication
```{r}

mat1 %*% mat2

```


Transposition
```{r}

t(mat1)

```



## List
In R, a list is a versatile and powerful data structure that can hold elements of different data types. Unlike vectors and matrices, lists can store heterogeneous data, such as numeric values, character strings, other lists, or even functions.

### Heterogeneous elements
Lists can contain elements of different data types. This flexibility allows you to organize and store diverse information together in a single data structure.


```{r}

list(25, "Elizabeth", c(1, 2, 3, 4), TRUE)

```


### Named Elements
List elements can be given names, making it easy to access and manipulate specific elements within the list.

```{r}

employee_list <- list(
  name = c("Elizabeth", "John"),
  age = c(15, 25),
  active_employee = c(TRUE, FALSE)
)

employee_list

```



### Nested List
Lists can contain other lists as elements, creating a nested data structure. This allows you to organize complex data hierarchically,
 
```{r}

list(
  name = "John",
  address = list(country = "Ghana", region = "Ashanti")
)

```


### Accessing Elements
In R, you can access elements in a list using different methods, depending on whether you want to access them by their index or their names.


```{r}

# first syntax using an index and double bracket `[[]]`
employee_list[[1]]


# second syntax using character names and double bracket
employee_list[["name"]]


# third syntax using list name and list-subsetting operator '$'
employee_list$name

```


```{r}
# access the first name in the list using the index

employee_list$name[1]

```

Multiple elements
```{r}
# access the multiple names in the list using the index

employee_list$name[c(1, 2)]

```


### Adding Elements to a list
To add new elements, we will use the ( $ ) operator to access the top element in the list and use the `c()` function to concatenate the new elements with the existing vector.

```{r}

# Update the name element

employee_list$name <- c(employee_list$name, "Mary")


# Update the age element
employee_list$age <- c(employee_list$age, 28)


# Update the active_employee element
employee_list$active_employee <- c(employee_list$active_employee, TRUE)


employee_list

```


### Deleting element
You can use the NULL value to indicate that the item should be removed. Setting an element in a list to NULL effectively deletes that particular element. 

All elements

```{r}

employee_list$active_employee <- NULL

employee_list

```


Deleting single elements using index
To delete single elements from a vector in a list, you can use indexing and subsetting.
```{r}

employee_list$name <- employee_list$name[-4]


employee_list$age <- employee_list$age[-4]

employee_list$gender <- employee_list$gender[-4]

employee_list

```



## Dataframe
In R, a dataframe is a two-dimensional data structure that is similar to a spreadsheet. It is one of the most commonly used data structures in R for data manipulation and analysis. Dataframes are especially well-suited for working with structured and tabular data.


```{r}

df <- data.frame(
  name = c("Elizabeth", "John"),
  age = c(15, 25),
  active_employee = c(TRUE, FALSE)
)

df

```


### access the column names using colnames


```{r}

colnames(df)

```


### Get the number of rows and columns

```{r}

# Number of rows using `nrow()` function
nrow(df)


# Number of columns using `ncol()` function
ncol(df)

 
# Or both rows and columns using `dim()` function
dim(df)

```


### structure

```{r}

str(df)

```


### adding new columns
To add a new column to a data frame, you can simply assign a vector of values to a new column name within the data frame.


```{r}

df$gender <- c("Female", "Male")

df

```


Or using `cbind()` function

```{r}

salary <- c(1500, 1500)

df <- cbind(df, salary)

df

```


### adding new rows
We can add new records (rows) using the `rbind()` function.


```{r}

df <- rbind(df, c("Mary", 28, TRUE, "Female", 1600))

df

```



## Accessing elements in a dataframe

Extract a column
```{r}

df$name

```

Using index row and columns to get specific records
```{r}

df[1, 1]  # The first row (1) and column (1) 

```


A single record and multiple columns
```{r}

df[1, ]

```


Multiple records and columns
```{r}

df[c(1, 2), c(2, 5)]  # Get the age and salary of each employee

```