# Logical Operators
Logical operators in R are used to combine or manipulate logical values (Boolean values: `TRUE` or `FALSE`) and are crucial for performing conditional evaluations, creating complex conditions, and controlling the flow of code execution.

R provides three main logical operators: `"AND" (&), "OR" (|), and "NOT" (!)`.


## And Operator ( & )
The AND operator evaluate two or more conditions and return `TRUE` if all conditions are true; otherwise, it return `FALSE`. It require that all individual expressions connected by '&' must be TRUE for final result to be `TRUE.`

```{r}

TRUE & TRUE       

FALSE & FALSE

1 & 1

```


```{r}

FALSE & TRUE

0 & 1

```



## Or Operator ( | )
The OR operator evaluate two or more conditions and return TRUE if at least one of the conditions is TRUE; otherwise, it returns FALSE. it requires that at least one of the individual expressions connected by `|` must be TRUE for the final result to be TRUE.

```{r}

TRUE | TRUE       

FALSE | FALSE

1 | 1

```


```{r}

TRUE | FALSE

1 | 0

```



## NOT Operator ( ! )
The NOT operator negates a single logical expression and return the opposite value. it converts `TRUE` to `FALSE` and `FALSE` to `TRUE.`

```{r}
!TRUE
!FALSE

!1
!0
```